# The backend abstraction: op dictionary and node registry {#sec-backend}


- The deepest internal work, deliberately last: replace stringly-typed
  TF operations with a backend-neutral op dictionary, and replace the
  doubly-linked R6 node objects with a flat registry of plain records.
  Grouping labels: `op-dictionary`, `node-registry`, and the internals
  half of `performance`.
- These are the enabling foundations for a second backend and for the
  2027 graph-rewriting features (milestone 12). They land last because
  they want the conformance suite (milestone 7) as their test oracle.

::: {.callout-warning}

## The `.internals` desync: a systemic argument for this milestone

Found independently in **two** extension packages, and the strongest
practical case for a stable internal API.

Extension packages reach into greta through `.internals$...`, and those
symbols are captured at **install** time. A greta upgrade therefore
silently desyncs every extension package until it is reinstalled -
nothing errors at load, and nothing tells the user.

The concrete instance: greta 0.6.0 renamed `tf_environment$batch_size`
to `.batch_size`. That single rename **broke greta.distributions' test
helpers**, and left **greta.dynamics' `batch_size` override silently
dormant** (greta.dynamics issue [#41](https://github.com/greta-dev/greta/issues/41)). Note the perverse consequence
there: fixing the *name* would **restore inert behaviour** - re-activate
an override that has quietly been doing nothing. Whoever takes that
issue must treat it as a behaviour change, not a typo fix.

Neither failure is a bug in the extension package. Both follow from
greta having no stable internal API - only `.internals`, a window onto
whatever the current implementation happens to be called. **The op
dictionary and node registry are what make a stable internal surface
possible**: once ops are recorded by backend-neutral name and nodes are
plain records, extension packages have a contract to depend on rather
than a snapshot of internals. Until then, the mitigation is a cross-repo
integration check that installs each extension package against greta's
development head.

(For a third instance of the same shape, see greta.checks: its
`.internals` import turned out to be unused copy-paste, which is the
only reason it survived 0.4.2 → 0.6.0 untouched.)
:::

**Why last, and why cheap now.** The classic reason to defer a core
rewrite is that it eats months of scarce maintainer attention. But a
mechanical, well-specified transformation guarded by a strong test
oracle is precisely the work that AI assistance makes cheap: the
maintainer cost becomes review hours, not writing hours. Nothing in
this year's user-facing plan strictly *needs* either rewrite - the
targets fix is targeted accessors, the unpin is install-layer, SNAPER
is sampler-layer - so they are scheduled deliberately, not out of
necessity.

**The op dictionary.** greta already stores each operation as a string
and resolves it late - `eval(parse(text = self$operation), envir =
self$tf_function_env)` at `R/node_types.R:167-171`, with the name
stored as a string at `R/node_types.R:120-122`. So greta is *already*
a late-bound, name-based design. The only problem is that the names
literally contain `tf$`, at ~96 call sites: `R/functions.R` (62),
`R/operators.R` (18), `R/extract_replace_combine.R` (9),
`R/transforms.R` (6), `R/chol2symm.R` (1). The move - the
`array-redesign.R` sketch in this repo, adopted as policy - is to
record ops by backend-neutral name and resolve them through a
per-backend table. A pure-R miniature of the whole contract runs
today:

```{r}
backend_r <- list(
  # 1. the ops: backend-neutral names, one implementation per backend
  ops = list(
    add    = `+`,
    matmul = `%*%`
  ),
  # 2. the backend owns compilation (a no-op for plain R;
  #    tf_function for TF; jit for JAX)
  compile = identity,
  # 3. the backend owns random-number discipline
  rng = list(
    make_seed  = \(seed) as.integer(seed),
    split_seed = \(seed) c(seed * 2L, seed * 2L + 1L)
  ),
  # 4. the backend owns dtype rules
  dtype = list(default_float = "double")
)

# a tiny "model" built against the contract, compiled by the backend
predict_fn <- backend_r$compile(
  \(x, beta) backend_r$ops$matmul(x, beta)
)

predict_fn(diag(2), matrix(c(0.5, 2)))
```

The four clauses - ops, compile, rng, dtype - are the definition of "a
greta backend"; the last three are what studying JAX taught us. The
work is mechanical and file-by-file: change one call site from
`op("add", ..., tf_operation = "tf$add")` to `op("add", ...,
operation = "add")`, and turn the single `eval(parse())` at
`R/node_types.R:167` into a table lookup. The 67 `tf_*()` shims in
`R/tf_functions.R` become the *values* of the TF backend's table -
they already are the per-backend implementations, just not yet indexed
by neutral names.

**The flat node registry.** Every greta array today is backed by an R6
node whose parent/child links are stored on *both* ends and mutated
together. This is the actual pair of methods - `add_parent()` appends to
`self$parents` and, in the same call, reaches into the parent and
appends `self` to *its* `children`:

```{r}
#| eval: false
# R/node_class.R:60-64 and 96-99 (the real methods, verbatim)
add_parent = function(node) {
  self$parents <- c(self$parents, node)
  node$add_child(self)
},
add_child = function(node) {
  self$children <- c(self$children, node)
},
```

The entries stored are the live R6 objects themselves, not ids, so after
`x$add_parent(y)` the object in `y$children` *is* `x` in the pointer
sense (`identical()` is true), and the whole connected component is
reachable from any node in it. Two consequences fall straight out of
this shape: node identity is the R6 environment address, which is why a
random-hex `unique_name` has to be minted at construction
(`create_unique_name()`, `R/node_class.R:262-264`; the collision hazard
is [#366](https://github.com/greta-dev/greta/issues/366)); and the
`children` list carries no information the parent lists don't - it is a
pure redundancy kept only to make downward traversal a field read. That
redundancy is exactly what Phase A below removes.

The registry replaces all of this with one flat table of plain records
referring to each other by id. Note the contrast with `add_parent()`:
`add_node()` writes **one** record, because there is no second `children`
field to keep in sync, and children are computed on demand instead:

```{r}
# a dag owns a single flat registry of nodes
new_dag <- function() {
  dag <- new.env()
  dag$nodes <- list()
  dag
}

new_node_id <- local({
  counter <- 0
  function() {
    counter <<- counter + 1
    sprintf("node_%03d", counter)
  }
})

add_node <- function(dag, type, parents = character(), value = NULL) {
  id <- new_node_id()
  # a node is plain data: no methods, no back-references, no pointers
  dag$nodes[[id]] <- list(
    id      = id,
    type    = type,
    parents = parents,
    value   = value
  )
  id
}

# children are derived from parents, not stored - contrast add_child()
children_of <- function(dag, id) {
  has_parent <- vapply(dag$nodes, \(node) id %in% node$parents, logical(1))
  names(dag$nodes)[has_parent]
}

# x is a variable; y = x * 2 becomes three plain records
dag <- new_dag()
x <- add_node(dag, "variable")
two <- add_node(dag, "data", value = 2)
y <- add_node(dag, "operation", parents = c(x, two))

children_of(dag, x)

# serialisation is trivially safe: the graph is a list of lists
identical(unserialize(serialize(dag$nodes, NULL)), dag$nodes)
```

| | R6 today | Flat registry |
|---|---|---|
| Edge storage | Doubly linked: `add_parent()` writes `self$parents` *and* the parent's `children` (`R/node_class.R:60-64`, `96-99`) | Single-ended: each record stores only `parents`; children derived (`children_of()`) |
| Node identity | R6 environment address; random-hex `unique_name` to compensate (`R/node_class.R:262-264`) | The record's id - value-addressable |
| Copying a model | Deep clone across the connected graph; currently errors ([#267](https://github.com/greta-dev/greta/issues/267)) - it recurses into the dag's `tf$Graph()` pointer (`R/dag_class.R:75`) *and* breaks shared-reference back-links | Plain list copy; no pointers on records, no shared identity to preserve |
| Saving/loading | greta arrays serialise safely; models/samplers hold Python handles patched by milestone 9 | The graph is a list of lists, safe by construction |
| Debugging | Stepping through `self$` chains in R6 environments | Plain `traceback()` through ordinary functions |
| Traversal cost | Recursion through R6 method calls (`parent_names(recursive = TRUE)`, `R/node_class.R:124-148`) | `vapply()` scan of a column; no per-node dispatch |

(The full code-grounded comparison, with every quote and `file:line`
checked against the source, is in `evidence/node-registry-comparison.md`.)

It is staged so it never blocks a release: (A) stop storing child
links, compute them from parents on demand - behaviour-preserving; (B)
parent references become ids resolved through the dag table - the
"linear register" of
[#247](https://github.com/greta-dev/greta/issues/247); (C) node
internals become plain records, after which the object-system question
(R6 vs S3 vs S7) shrinks to a small, local, reversible choice. (On S7:
not fit for the mutable core today - even ggplot2 kept its mutable
layer out of S7 - but fine for user-facing value classes later.)


## The speed family (evidence for the performance cluster)

- [#294](https://github.com/greta-dev/greta/issues/294) reports
  "There is a huge efficiency gain to be had from setting the number
  of chains to a large number in greta"; the benchmark: 4 chains →
  185 effective samples/second; 1024 chains → 12,980.
- [#547](https://github.com/greta-dev/greta/issues/547) - "during
  warmup we step between R and TF every 4 iterations... in TF2, we
  can avoid doing this on the R side and do it completely in TF."
- [#358](https://github.com/greta-dev/greta/issues/358) - draws
  serialisation "accounted for most of the pain" on big models.
- [#210](https://github.com/greta-dev/greta/issues/210) - loops "very
  slow to define and execute" (model-definition overhead; a
  node-registry trigger).
- [#309](https://github.com/greta-dev/greta/issues/309) - subsetting
  shims "don't always use the most efficient operations".



## Background reading for the backend decision

These notes were written with the TensorFlow unpin, on the reasoning
that the unpin was where the backend question would first bite. It
shipped without needing them, so they sit here instead,
with the milestone that actually has to answer the question.

### JAX in one paragraph, for R developers

JAX is "NumPy plus function transforms": write ordinary functions over
arrays, then ask for transformed versions - `grad(f)` (the gradient
function), `jit(f)` (compiled via XLA), `vmap(f)` (vectorised). No
variables, no graph state, arrays immutable, randomness via explicit
split keys. `jax.jit` and `tf.function` are close cousins (trace once,
cache compiled), which is why greta's architecture translates: the dag
trace is already effectively a pure `free_state → log_prob` function.
R reaches JAX through reticulate only, and the precedent is proven -
Posit's keras3 drives the JAX backend from R in production. The three
R-bridge adaptations greta would need: pure-function discipline in the
dag (mostly already true), a key-splitting step per sampler burst, and
retrace-safe burst lengths (greta solved the same problem for TF).


### What other PPLs teach

- **Stan**: owns everything down to C++ autodiff - zero framework
  risk, unsustainable maintenance for two people; *usable* via
  BridgeStan (compiled model exposes `log_density_gradient()` to R).
- **PyMC**: the only PPL to survive backend death, twice
  (Theano → Aesara → PyTensor) - survived *because* it owned its graph
  and its backend was forkable pure Python. The direct analogue of
  greta's position.
- **NumPyro**: all-in on JAX, thriving; hedge is a small portable
  codebase.
- **Edward**: coupled to TF1 internals; did not survive the TF2
  transition.

The survivor pattern: own the model representation; ask the backend
only for tensors and gradients through a thin, enumerable boundary.
greta already has this shape.


### Modern R concurrency: mirai and mori

**mirai** (r-lib, v2.7.x) is the async engine under targets' crew,
purrr's parallel map, and Shiny. For greta: `future.mirai` may work
with greta's chains unchanged today (hypothesis to verify - an October
task); **persistent daemons** would let Python/TF initialise once per
worker and be reused across `mcmc()` calls - a substantial
parallel-chains win; and its `serial_config()` refhooks are the right
interception point if pointer-*recipes* ever need to cross processes
(they complement, not replace, the session-portability workstream's
self-healing design).
**mori** (v0.2.x, same author) offers zero-copy shared memory for
plain R data across local workers - nice-to-have for sharing large
input data between chains, irrelevant to the Python-pointer problem,
too young to plan around.


### The mrc-ide stack: odin2, monty, dust2

Imperial's infectious-disease group rebuilt their ODE/state-space
stack: **odin2** (R-like DSL transpiling to C++), **dust2** (the C++
engine, particle filters), **monty** (composable models + samplers in
R). Verdict for greta: **complementary, not competitive** - they own
sequential Monte Carlo over dynamical systems; greta owns general
Bayesian graphical models with automatic gradients. Two lessons
matter: monty's `monty_model(density, gradient)` interface is exactly
the shape of greta's sampler-interface workstream (external
validation, and a feasible interop experiment); and their gradient
story - symbolic differentiation of a
restricted DSL, or compiled C++ AD, never general R-level autodiff -
shows what an R-native greta backend would honestly require: symbolic
reverse-mode differentiation of greta's own DAG. That is the credible
"greta without Python" route, and monty's differentiable DSL is its
working small-scale precedent.


### Object systems: the S7 verdict

S7 (v0.2.x) is value-semantics by design; the OOP working group's own
scoping excludes mutable objects; environment subclassing was removed
pending redesign; and ggplot2's 4.0 migration - the flagship - kept
its mutable ggproto layer *out* of S7. No large R6→S7 migration
precedent exists. Conclusion: S7 is not a candidate for greta's node
core today; it remains a reasonable future choice for user-facing
value classes (the vctrs work at the edges). The genuine core contest
is R6 (status quo) versus S3-over-the-registry, and the node-registry
workstream makes that choice small, local, and reversible - which is
why it need not be decided now.


## Backend evidence: the empirical tests behind the unpin {#sec-evidence}

This section records the empirical evidence for the unpin and the wider
backend strategy: the two claims below were tested by actually running
them (July 2026), and the unpin and every later backend claim rest on
these results.

::: {.callout-note}

### The short version

We tested the two most important backend claims by actually running
them (July 2026): **latest TensorFlow works with greta's frozen
probability library** (so the version pin can be lifted), and **the
same probability library runs on JAX** (so greta has a credible second
backend). This appendix records those tests, the health of each
upstream project, and the contingency ladder if things get worse.
:::


### The empirical tests

Both tests were run in fresh uv-provisioned Python 3.12 environments,
exercising the *exact* surface greta uses - enumerated by grepping
greta's `R/` directory: ~25 `tfp$distributions$*` classes, the
bijectors (Chain, Exp, Shift, Reshape, Transpose,
CholeskyOuterProduct, FillScaleTriL), `tfp$mcmc$` (HMC, RWM, Slice,
`sample_chain`), `tfp$optimizer$` (BFGS, Nelder-Mead), and
`tfp$math$value_and_gradient`. The one Keras surface greta does touch -
its 8 legacy optimisers, `tf$keras$optimizers$legacy$*`
(`R/optimisers.R:359-625`) - is *not* exercised by these tests, and is
the one part of the unpin still to be checked under TF 2.21 + tf-keras
(see the caveat on Test 1).

**Test 1 - the unpin.** `tensorflow==2.21.0` +
`tensorflow-probability==0.25.0` + `tf-keras==2.21.0` (numpy 2.5.1):
distributions including WishartTriL, CholeskyLKJ,
DirichletMultinomial; bijector chains; gradients; a 300-step HMC chain
inside `tf.function`; `bfgs_minimize`. **All passed** - three TF minor
versions past TFP 0.25's officially tested pairing. Caveat: this
exercises the TFP surface and `bfgs_minimize`, but not greta's 8 legacy
Keras optimisers (`tf$keras$optimizers$legacy$*`, `R/optimisers.R:359-625`);
because the Keras 2/3 break is exactly what `tf-keras` exists to
paper over, those optimisers should be run under this stack before the
unpin ships.

**Test 2 - the JAX substrate.** The same kernel surface through
`tfp.substrates.jax`, at both ends: TFP 0.25 with JAX 0.4.35 (the
frozen-but-coherent pairing) and tfp-nightly with JAX 0.10.2 (the
living pairing). **Both passed**, including HMC under `jax.jit`. One
config note: greta's float64 default needs `jax_enable_x64`.

Key mechanics behind Test 1: TF 2.16 made Keras 3 the default and
broke TFP at import time
([tensorflow/probability#1774](https://github.com/tensorflow/probability/issues/1774)).
The fix is the `tf-keras` package (Keras 2 repackaged), which is
released in lockstep with every TF release - through 2.21.0 (March
2026) - and `TF_USE_LEGACY_KERAS=1` set before Python initialises.
TFP 0.25 declares no TF upper bound, so it resolves cleanly.


### Upstream health, grounded in dates

| Project | Status (July 2026) | Evidence |
|---|---|---|
| TFP | Frozen releases, live repo | Last release 0.25 (Nov 2024); ~weekly maintenance commits through July 2026, incl. JAX-team members; nightly's JAX substrate tracks latest JAX while released 0.25's does not |
| TF | Slowing, years from dead | 2.21 (Mar 2026); cadence now ~2/year; TensorBoard dep dropped; Google migrating internal models TF→JAX with dedicated tooling |
| tf-keras | Healthy escape hatch | Lockstep release with every TF version since Mar 2024 |
| JAX | Very much alive | Gemini/DeepMind existential dependency; quarterly-ish releases; small documented breaking changes per release |
| torch (R) | Healthy, no Python | Posit/mlverse maintained, CRAN; ~8 distributions, no bijectors/kernels - the PPL layer would be greta's to write |

The asymmetry that matters: **TFP is pure Python (~6.7 MB) - forkable
by a two-person team. TF is C++ - not forkable.** greta's exposure is
the TFP contract, and it is short.


### The contingency ladder

| Rung | Scenario | Response | Effort |
|---|---|---|---|
| 0 | TFP stalled but working (today) | tf-keras unpin + `tf < 2.22` cap + CI canary | days - **doing now** |
| 1 | TFP breaks on future TF/Python, no release comes | Freeze a known-good uv lockfile; vendor/patch affected TFP modules (pure Python) | days-weeks |
| 2 | TF terminal decline, TFP-at-HEAD stays JAX-alive | JAX substrate backend via the op dictionary | months (2-4) |
| 3 | Google archives TFP entirely | Micro-fork the ~30 TFP files greta touches, and/or move to Distrax + BlackJAX | months |
| 4 | Python exit demanded | Stan transpile via BridgeStan (months) / torch-R port (6-12 mo) / own densities + pluggable AD (year+) | see left |

Preparation that cheapens every rung (all in the 2026 plan): the
conformance suite, stateless seeds, op de-stringing, and composing new
distributions from greta ops where cheap.



## Supporting multiple backends in greta?

This is the one strategic question deliberately left open, and outside
input would be especially valuable here.

**The situation.** TFP, greta's probability library, is in maintenance
mode. History gives a clear steer: modelling languages coupled tightly
to their computational engine tend not to survive that engine's death
- Edward, built on TF1 internals, did not survive the TF2 transition.
Languages that own their model representation and treat the engine as
swappable tend to survive - PyMC has outlived two engine deaths
(Theano, then Aesara). greta already has the survivable shape: its
real dependency is not "TensorFlow" but a short, enumerable list of
TFP functions - roughly 25 log-densities, 8 bijectors, 3 MCMC kernels,
2 optimisers - and that list is what any endgame needs to replace or
keep working. The full analysis is in @sec-evidence.

**The candidate endgames:**

| | Endgame | What it means | Effort | Python? |
|---|---|---|---|---|
| (a) | Stay in TFP-world | Unpin now; JAX substrate as second backend | days now, months later | yes |
| (b) | torch-R flagship | Write ~25 log-densities, bijectors, kernels against torch's R-native autodiff | 6-12 months | **no** |
| (c) | Own the densities | Distributions become compositions of greta ops; backends are interchangeable gradient providers | year+ | optional |
| (d) | Stan transpile | Generate Stan code from the dag; gradients via BridgeStan | months | no (C++ toolchain) |

```{r}
#| eval: false
# what a user might eventually see, whichever way this goes:
# a session-level backend choice, in the spirit of future::plan()
# and keras3::use_backend()
greta_backend(backend = "jax")
m <- model(y)
draws <- mcmc(m, n_samples = 1000)
```

**Why it can stay open cheaply.** The endgame-neutral preparation -
the conformance suite (milestone 7), stateless seeds (milestone 4),
op de-stringing (milestone 11) - is identical under all four endgames.
The one day-to-day choice it *does* affect: when adding a new
distribution ([#369](https://github.com/greta-dev/greta/issues/369)), delegate to `tfp$distributions$X`, or compose it
from greta ops (portable to every future backend for free)? Default
until decided: compose where cheap, delegate where not. Issue
[#586](https://github.com/greta-dev/greta/issues/586) is the tracking
home for whichever way this resolves.



### [Issue #155](https://github.com/greta-dev/greta/issues/155)

- 2018-04-26
- (none)
- Cache the results of operations on data greta arrays.
- problem: constant sub-expressions recompute each call.
- finding: constant-folding is natural once ops are addressable by name in the op table.
- fix: cache constant op results at define time.
- priority: low-medium.

### [Issue #156](https://github.com/greta-dev/greta/issues/156)

- 2018-04-27
- (none)
- Add types to greta arrays.
- finding: types (continuous/discrete, shape, dtype) become a field on the plain node record.
- fix: carry a type field in the registry; use it for validation and the dtype clause.
- priority: medium (enabler for milestone 9 and better errors).

### [Issue #247](https://github.com/greta-dev/greta/issues/247)

- 2018-11-08
- (none)
- Generalise greta's DAG construction - the "linear register of already-defined nodes".
- finding: this is the design the flat registry implements; it also settles greta keeping its own shape semantics at op-recording time (avoiding the broadcasting ambiguity raised on the thread).
- fix: land Phases A-C of the registry.
- priority: medium (the design spine of this milestone).

### [Issue #309](https://github.com/greta-dev/greta/issues/309)

- 2019-08-17
- (none)
- Speed up `tf_extract()`/`tf_replace()`.
- finding: the subsetting shims "don't always use the most efficient operations" (performance family).
- fix: rewrite the extract/replace ops against the op table with efficient primitives.
- priority: low-medium.

### [Issue #581](https://github.com/greta-dev/greta/issues/581)

- 2022-12-20
- Up Next
- Implement `outer()` in greta.
- finding: a new op best added once the op dictionary is the place new ops go.
- fix: add `outer` as a neutral-named op with a TF implementation.
- priority: low.

### [Issue #586](https://github.com/greta-dev/greta/issues/586)

- 2023-04-20
- (none)
- Alternative backends (torch, xtensor).
- finding: this is what the op dictionary exists to enable; it also connects to the backend-endgame fork below.
- fix: once the op table lands, a second backend is a new table of implementations.
- priority: medium (2027; strategy-dependent).

### [Issue #705](https://github.com/greta-dev/greta/issues/705)

- 2024-08-16
- (none)
- Method dispatch to avoid `if (class(thing)) ... else ...`.
- finding: the manual class-branching is exactly what plain records + proper dispatch remove.
- fix: replace class-branches with S3/S7 dispatch as the registry lands.
- priority: low-medium.


