# Faster and adaptive sampling {#sec-adaptive-sampling}

- The sampling *capability* release: land the adaptive HMC sampler,
  give the log-prob function a data interface, and name the
  `(log_prob, gradient)` boundary that samplers actually depend on.
  Grouping labels: the `mcmc` cluster, `sampler-interface`, and the
  sampling half of `performance`.
- These issues cohere around one object - the sampler - and its
  contract with the model. The sampling *bugs* are not here: they ship
  earlier and separately, in milestone 2.

**The sampler interface.** A sampler needs exactly two things from a
model: the log-probability of a parameter vector, and its gradient.
greta's dag *already computes the value* at
`R/dag_class.R:61-66` (`tf_log_prob_function_adjusted/_unadjusted`), and
the sampler's gradient is autodiffed by the TFP kernel directly from
that log-prob function (greta already does tape-based autodiff for the
optimiser's Hessian - `tf$GradientTape` in `calculate_one_hessian` at
`R/dag_class.R:497-519`). Naming
this pair the official boundary means samplers stop caring what
computed them. The demonstration below is a complete HMC sampler
written against nothing but that closure, with no reference to
TensorFlow:

```{r}
# a model, from the sampler's point of view: value and gradient only
# (standard bivariate normal, standing in for greta's dag machinery)
lp <- list(
  value    = \(theta) -0.5 * sum(theta^2),
  gradient = \(theta) -theta
)

# a complete HMC sampler against the closure interface
sample_hmc <- function(lp, n_samples, theta, eps = 0.1, steps = 10) {
  draws <- matrix(NA_real_, n_samples, length(theta))
  for (i in seq_len(n_samples)) {
    p0 <- rnorm(length(theta))
    p <- p0
    q <- theta
    # leapfrog integration, using only lp$gradient()
    p <- p + eps / 2 * lp$gradient(q)
    for (s in seq_len(steps)) {
      q <- q + eps * p
      last_step <- s == steps
      p <- p + (if (last_step) eps / 2 else eps) * lp$gradient(q)
    }
    # accept/reject, using only lp$value()
    log_accept <-
      lp$value(q) - lp$value(theta) - 0.5 * sum(p^2) + 0.5 * sum(p0^2)
    if (log(runif(1)) < log_accept) {
      theta <- q
    }
    draws[i, ] <- theta
  }
  draws
}

set.seed(2026)
draws <- sample_hmc(lp, n_samples = 2000, theta = c(0, 0))

# recovers the standard normal
round(colMeans(draws), 2)
round(apply(draws, 2, sd), 2)
```

There is external validation: monty - the sampler layer from
Imperial's mrc-ide group - uses *exactly* this `(density, gradient)`
interface, so a greta model could be wrapped as a `monty_model` as a
cheap interop experiment.

**The tension, resolved by layering.**
[#547](https://github.com/greta-dev/greta/issues/547) pulls the other
way: for speed, push the whole chain - warmup tuning included - into
the backend. Both halves already exist. The fused path is the status
quo: greta already runs `tfp$mcmc$sample_chain()`
(`R/sampler_class.R:508`) inside a compiled `tf_function`
(`define_tf_evaluate_sample_batch()`, `R/sampler_class.R:74`). The
closure is the portability contract; backends may *additionally* offer
a fused fast-path. SNAPER lands against the fused path first (it is TFP
code today); [#547](https://github.com/greta-dev/greta/issues/547) later is a change *within* that path (construct
TFP's self-adapting kernel instead of stepping adaptation from R every
few iterations), not a rearchitecture.


## Sprint item: TFP sampler adaptation with progress-bar updating

- not a GitHub issue, and **deliberately not filed as one**: it is
  implementation detail for [#547](https://github.com/greta-dev/greta/issues/547)
  and [#765](https://github.com/greta-dev/greta/issues/765) rather than
  standalone work, so it is left until this milestone is actually picked up.
  From the [#322](https://github.com/greta-dev/greta/issues/322) sprint plan
  (§2.3, "Use TFP sampler adaptation (with progress bar updating)").
- Let TensorFlow Probability adapt the sampler's step size (and mass
  matrix) inside the compiled chain, instead of stepping adaptation
  from R between short bursts, and drive the progress bar from the
  chain's traced diagnostics.
- problem: greta adapts in R. `tune_epsilon()`
  (`R/sampler_class.R:411-449`) is a hand-coded dual-averaging update
  and `tune_diag_sd()` (`R/sampler_class.R:450-473`) shrinks the mass
  matrix, both invoked by `tune()` (`R/sampler_class.R:407-409`)
  between short `run_burst()` calls (`R/sampler_class.R:207-217`).
  Every burst hands control back to R - exactly the R<->TF
  round-tripping that [#547](https://github.com/greta-dev/greta/issues/547)
  wants to remove.
- finding: TFP already provides this. Build the leapfrog/HMC kernel
  with `store_parameters_in_results = TRUE` (so a wrapper may rewrite
  `step_size` on the fly), then wrap it in
  `tfp$mcmc$DualAveragingStepSizeAdaptation` (it matches greta's
  current dual-averaging scheme; `SimpleStepSizeAdaptation` is the
  simpler alternative) with `num_adaptation_steps` set to the warmup
  length. `sample_chain()` (`R/sampler_class.R:508`) then runs the
  whole warmup-plus-sampling run inside the existing `tf_function`,
  with TFP updating the step size internally each iteration.
- progress bar: the bar hooks through `trace_fn`
  (`R/sampler_class.R:512`), which today returns the entire
  `kernel_results`. Narrow it to just what the bar needs - the
  adapted step size (`new_step_size`) and
  `inner_results.log_accept_ratio` (acceptance is
  `exp(min(log_accept_ratio, 0))`) - and surface those per burst (or
  via a `py_func` callback) so the bar reflects live acceptance and
  step size. This also feeds
  [#773](https://github.com/greta-dev/greta/issues/773) (abstract the
  progress bars).
- fix: construct the adapting wrapper in `define_tf_kernel()`; set
  `store_parameters_in_results = TRUE`; replace the R-side
  `tune()`/`tune_epsilon()`/`tune_diag_sd()` stepping with one
  warmup-length `sample_chain()` under the wrapper; drive the progress
  bar from a trimmed `trace_fn`. This is the same fused-path move
  [#547](https://github.com/greta-dev/greta/issues/547) describes, and the mechanism the SNAPER work ([#765](https://github.com/greta-dev/greta/issues/765)/#779)
  already relies on.
- priority: medium (unlocks [#547](https://github.com/greta-dev/greta/issues/547); groundwork for the adaptive sampler).


::: {.callout-note collapse="true"}
## Aside: the Geweke check harness for SNAPER

The in-progress Geweke testing code for `adaptive_hmc()` (from the
local `inst/greta-geweke-ahmc.R` work accompanying
[#779](https://github.com/greta-dev/greta/pull/779)):

```{r}
#| eval: false
devtools::load_all(".")

# parameters ----
n <- 10
mu1 <- 0
sd1 <- 2
sd2 <- 1

# prior (n draws) ----
p_theta <- function(n) {
  rnorm(n, mu1, sd1)
}

# likelihood ----
p_x_bar_theta <- function(theta) {
  rnorm(n, theta, sd2)
}

# define the greta model ----
x <- as_data(rep(0, n))
greta_theta <- normal(mu1, sd1)
distribution(x) <- normal(greta_theta, sd2)
model <- model(greta_theta)

# mcmc parameters ----
n_iter <- 250
n_warmup <- 10000
n_chains <- 20
n_thin <- 5
geweke_sampler <- adaptive_hmc()

#  checking ----
time_taken <- system.time({
  geweke_ahmc <- check_geweke(
    sampler = geweke_sampler,
    model = model,
    data = x,
    p_theta = p_theta,
    p_x_bar_theta = p_x_bar_theta,
    chains = n_chains,
    niter = n_iter,
    warmup = n_warmup,
    thin = n_thin
  )
})

# processing ----
geweke_thin <- apply_thinning(geweke_ahmc, n_thin)

geweke_stat <- geweke_ks(geweke_thin)

geweke_stat

the_time <- round(time_taken[3])

cat("This took", the_time, "seconds")

# ---- plotting ----
plot_metadata <- build_plot_metadata(
  sampler = geweke_sampler,
  time = the_time,
  n_warmup = n_warmup,
  n_iter = n_iter,
  n_chains = n_chains,
  n_thin = n_thin,
  ks_stat = geweke_stat
)

png(plot_metadata$qq_filename, width = 1200, height = 800, res = 150)

geweke_qq(geweke_thin, title = plot_metadata$qq_title)
dev.off()

png(plot_metadata$coda_filename, width = 1200, height = 800, res = 150)
plot(geweke_ahmc$draws)
dev.off()

bplot_dens <- bayesplot::mcmc_dens_chains(geweke_ahmc$draws)

ggplot2::ggsave(
  filename = plot_metadata$bayesplot_filename,
  plot = bplot_dens,
  width = 1200,
  height = 800,
  dpi = 150,
  units = "px",
  bg = "white"
)

```
:::


## [Issue #181](https://github.com/greta-dev/greta/issues/181)

- 2018-06-01
- documentation
- Reusing one model across different datasets.
- finding: the mechanism is the data interface ([#739](https://github.com/greta-dev/greta/issues/739)).
- fix: document reuse once [#739](https://github.com/greta-dev/greta/issues/739) lands.
- priority: low (follows [#739](https://github.com/greta-dev/greta/issues/739)).

## [Issue #314](https://github.com/greta-dev/greta/issues/314)

- 2019-08-20
- (none)
- "greta goes NUTS" - a NUTS sampler.
- finding: the `nuts` branch is TF1-era; TFP provides NUTS today.
- fix: expose TFP NUTS through the sampler interface; delete the old branch.
- priority: medium.

## [Issue #358](https://github.com/greta-dev/greta/issues/358)

- 2020-12-15
- (none)
- Save draws to disk from TF directly rather than serialise into R.
- finding: draws serialisation "accounted for most of the pain" on big models (performance family).
- fix: stream draws to disk from the fused path.
- priority: medium.

## [Issue #547](https://github.com/greta-dev/greta/issues/547)

- 2022-08-09
- TF2, impact-1
- Do more in TensorFlow instead of stepping back to R each burst.
- finding, code-verified: greta steps R↔TF every few warmup iterations; the fused path already exists (`R/sampler_class.R:74,509`).
- finding, measured: the fused path traces **once** per sampler - `tf_evaluate_sample_batch` is a real `tf.function` and `experimental_get_tracing_count()` returns 1 after 200 iterations, after 1000, and with 4 chains. Being traced is therefore *not* what is costing anything; the round-trips are. Measured bursts for one chain, quiet: warmup 500 + 500 samples → 168; warmup 500 + 50 samples → 168; warmup 50 + 500 samples → 18. The count tracks **warmup only** - the whole sampling phase is a single burst - so it is roughly one return to R every 3 warmup iterations, matching the issue's "every 4 iterations". `one_by_one = TRUE` gives one burst per iteration.
- finding: what actually drags R back in per burst is tuning (`update_welford()`, `tune()`), redrawing `sampler_parameter_values()`, and progress reporting - not the sampling itself.
- finding: progress reporting for parallel chains **already works by watching files**. `create_log_file()` (`R/utils.R:107`) hands each sampler three tempfiles; the sampler `writeLines()` percentage/progress into them (`R/inference_class.R:77-82`, `record()` at `R/utils.R:130`) and the parent session polls them back with `readLines()`/`read.table()` (`R/callbacks.R:4-20`).
- direction (Nick, 2026-08-20): explore **mirai** and **mori** for that file-watching layer, so progress can be observed without the sampler returning to R to report it. Caveats to check first: **mori is shared memory, not a file watcher** - "Shared Memory for R Objects", Charlie Gao, v0.2.2 (July 2026), POSIX shm/Win32 file mapping via ALTREP - so it would *replace* the tempfiles rather than watch them; and it needs **R >= 4.3**, above greta's current `R (>= 4.2.0)`, so it cannot be a hard dependency without a bump. See the concurrency notes in the backend chapter.
- fix: build TFP's self-adapting kernel; run longer bursts - a swap inside the fused function. Decoupling progress reporting is the complementary half: longer bursts are only usable if the user can still see progress.
- priority: medium (speed; 2027).

## [Issue #568](https://github.com/greta-dev/greta/issues/568)

- 2022-11-04
- TF2, impact-1
- Implement new TFP optimisers.
- finding: greta's optimiser list predates current TFP; overlaps [#244](https://github.com/greta-dev/greta/issues/244)/#569.
- fix: wrap the current TFP optimisers; refresh `R/optimisers.R`.
- priority: medium.

## [Issue #569](https://github.com/greta-dev/greta/issues/569)

- 2022-11-04
- Up Next, TF2
- Extract convergence and iteration info from TF2 optimisers.
- finding: TFP optimisers expose convergence state greta discards.
- fix: surface convergence/iterations on the optimiser result.
- priority: medium.

## [Issue #739](https://github.com/greta-dev/greta/issues/739)

- 2024-11-01
- (none)
- Add a data interface into the log-prob function.
- problem: no clean way to swap data without rebuilding the model - the enabling hook for reuse ([#181](https://github.com/greta-dev/greta/issues/181)) and info criteria ([#519](https://github.com/greta-dev/greta/issues/519)).
- finding: the sampler-interface closure (milestone 6) is the natural place to parameterise by data.
- fix: parameterise the compiled log-prob closure by its (non-constant) data nodes so data can be swapped without a rebuild - turn swappable data into arguments of the traced function in `define_tf_log_prob_function()`/`generate_log_prob_function()` (`R/dag_class.R:50-56`), driven by the data-node "using constants" flag (`R/node_types.R`). Dependency inversion: this milestone-3 issue is naturally built on the milestone-6 sampler-interface closure - schedule after M6, or pull just the data hook forward if [#519](https://github.com/greta-dev/greta/issues/519) is wanted sooner.
- **warning for whoever picks this up:** `dag_class` exposes `get_tf_data_list()`/`set_tf_data_list()` (`R/dag_class.R:599-606`), which look like an existing data-swap seam. **They are not.** Tested directly: the data list is empty and swapping through it does not move the log prob (`-13.97077` before and after). Data is baked in at trace time either way - as `tf$constant` under the constants branch, as an eager `as_tensor` otherwise (`R/node_types.R:42-68`). The source comment concedes it: *"we might not need this anymore as it was to do with stashing things for use in the feed_dict later"*. It is **vestigial TF1 feed_dict plumbing** to be removed or repurposed, not built on. It in turn blocks [#519](https://github.com/greta-dev/greta/issues/519) and enables [#181](https://github.com/greta-dev/greta/issues/181).
- proposed fix: see [fixes/issue-739.md](fixes/issue-739.md).
- priority: medium (enabler).

## [Issue #765](https://github.com/greta-dev/greta/issues/765)

- 2025-02-12
- (none)
- Add the SNAPER HMC sampler.
- finding: PR [#779](https://github.com/greta-dev/greta/pull/779) implements `adaptive_hmc()`; the blocker is *accuracy, not code* - it must pass the Geweke check (harness in the aside below).
- fix: rebase [#779](https://github.com/greta-dev/greta/issues/779), make Geweke pass, land against the fused path.
- priority: high (headline).

## [Issue #790](https://github.com/greta-dev/greta/issues/790)

- 2025-05-13
- (none)
- Code comparing HMC to adaptive HMC.
- finding: the efficiency case - `adaptive_hmc()` far outperforms `hmc()` on effective sample size.
- fix: fold the comparison into the sampler docs and the acceptance evidence.
- priority: medium (evidence for [#765](https://github.com/greta-dev/greta/issues/765)).


